Environmental and geographic variables are effective surrogates for genetic variation in conservation planning

### 2016-10-11 code/rmarkdown/article.Rmd
+ article within page pnas limits

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Hugh P. Possingham
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  All authors developed the study and drafted the manuscript. JOH performed all analyses.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for processes to continue. However, evolutionary processes are not often explicitly considered when deciding where to site protected areas because genetic data require considerable resources to obtain and analyse. Here we show that environmental and geographic distance variables are effective surrogates for adaptive and neutral intra-specific variation for conservation planning, and that these surrogates can be used to deliver more effective solutions. We obtained distribution and genetic data from the international IntraBioDiv project for 27 species collected over the European Alps. Samples were collected using a gridded sampling scheme. We used these cells as planning units to generate solutions. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the planning units. Our first analysis confirms that the the spatial patterns of adaptive and neutral genetic variation correlate with those of environmental and geographic variation (respectively). Our second analysis verifies that solutions that secure more adaptive genetic variation also tend to secure more adaptive variation, and similar patterns were found for neutral genetic variation and geographic variables. Our third analysis found that solutions generated using additional targets to explicitly secure a representative sample of the environmental and geographic variation across a species range resulted in solutions that secure more neutral genetic variation for most species. Additionally, we found that most solutions--even those generated without surrogate or genetic data--tended to secure most of the adaptive genetic variation for most species. Overall, our analysis provides a rigorous examination of two potential surrogates for genetic variation. 
significance: |
  Protected areas are used to preserve biodiversity patterns and processes. Since the resources available for conservation are limited, protected areas need to be sited in cost-effective locations. However, despite the stated emphasis on biodiversity processes, evolutionary processes are not often used to guide reserve selection. This is due to the fact that genetic data is often too costly to obtain and process. Here we found that environmental and geographic variables are cost-effective surrogates for genetic data. These variable can be used to deliver candidate reserve systems that secure the adaptive and neutral evolutionary processes for a greater number of species with minimal additional cost. This research highlights the importance of explicitly considering biodiversity processes during conservation planning exercises.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Excellence for Environmental Decisions (CEED) and the Landscape Ecology and Conservation Group (LEC) at The University of Queensland.
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

<!-- Introduction -->
Protected areas spearhead conservation efforts [@r440]. They buffer species from anthropogenic impacts, providing places for species to persist. Since the resources available for conservation are limited, protected areas need to be sited in places that fulfill conservation objectives for minimal cost [@r255]. To achieve this, conservation planning exercises often generate plans for entire networks of protected areas (prioritisations) to allocate resources cost-effectively and preserve broad-scale biodiversity processes. Intra-specific genetic variation is fundamental to long-term species persistence [@r165], and as a consequence, there has been increasing interest in designing prioritisations that capture and cultivate this aspect of biodiversity [@r67; @r3; @r61; @r499].

Broadly speaking, though the strength of natural selection is a continuous force, genetic variation can be classified as either adaptive or neutral [reviewed in @r480]. Adaptive genetic variation is associated with loci that significantly affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals [eg. @r456; @r487; @r480; @r490], though phenotypic-genetic associations are more likely to identify functionally important genomic regions for individual species [eg. @r500]. If they represent the full range of adaptive variation, protected areas can help ensure that populations with particularly beneficial adaptations are not lost [@r3; @r43]. In contrast, neutral genetic variation is associated with loci that do not significantly affect fitness, but instead reflects the evolutionary history of different populations. By representing the full range of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity [@rXXX]. Thus optimally sited protected area networks would aim to represent both adaptive and neutral patterns of genetic variation for species of conservation interest [@r67; @r3; @r43].

Recently, the use of surrogate data has been proposed to generate prioritisations that adequately preserve intra-specific genetic variation--without needing to directly utilise genetic data to achieve this. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been proposed to guide reserve selection as surrogates for adaptive genetic variation [eg. environmental conditions; @r202; @r247; @r2]. By preserving a more representative sample of environmental conditions across the species' ranges, prioritisations could be generated that also capture more adaptive variation. However, the effectiveness of environmental surrogates in representing adaptive genetic variation remains unverified. Neutral genetic variation arises from a reduction in gene flow between individuals. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood [@r501], and potential surrogates for neutral variation have been based on variables that predict the levels of connectivity between planning units. For instance, isolation by distance [IBD; @r259] predicts that populations further apart will be more weakly connected, exchange less genetic material, and in turn have less genetic variation in common. Based on this idea, geographic surrogates based on distance have been found to improve conservation prioritisations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain or habitat type; @r474; @r497]--the typical situations in most conservation planning exercises [eg. @r197].

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods to detect loci under selection, and they returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions was needed to sufficiently describe their patterns of adaptive (all $K = `r unique(na.omit(adaptive.k))`$; Table S2) and neutral genetic variation (minimum $K = `r min(neutral.k)`$; mean $K = `r round(mean(neutral.k),2)`$; maximum $K = `r max(neutral.k)`$; Table S2). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S29--S54) generally showed substiantial spatial auto-correlation (eg. _Cerastium uniflorum_, Figure S34; _Dryas octopetala_, Figures S36), suggesting that the environmental and geographic variables may prove to be effective surrogates.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental and geographic variation correlated with the broad-scale spatial patterns in adaptive and neutral variation for most species (respectively). We constructed dissimilarity matrices between the planning units. For each species, we constructed matrices expressing the differences between planning units as a function of their environmental characteristics and geographic position, and the adaptive (if detected) and neutral genetic characteristics between individuals. The spatial patterns of environmental variation was significantly correlated with that of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models). Geographic distance between planning units also correlated significantly with the spatial pattern of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these patterns, our next step was to determine whether these variables are effective surrogates for conservation planning.
```{r, echo=FALSE, results='asis'}
curr.file <- tempfile(fileext='.tex')
x <- correlation.model.summary.DF %>%
	select(-contains('raw')) %>%
	mutate(
		Environmental.corr.P=pvalString(Environmental.corr.P),
		Geographic.corr.P=pvalString(Geographic.corr.P),
		Geographic.Marginal.R2=pvalString(Geographic.Marginal.R2),
		Geographic.Conditional.R2=pvalString(Geographic.Conditional.R2),
		Environmental.Marginal.R2=pvalString(Environmental.Marginal.R2),
		Environmental.Conditional.R2=pvalString(Environmental.Conditional.R2),
		blank=rep('', length(Species)),
		Species=paste0('\\textit{', gsub('\\ ', ' ', Species), '}')) %>%
	mutate(Environmental.Chisq=sapply(round(Environmental.Chisq,2), as.character)) %>%
	mutate(Geographic.Chisq=sapply(round(Geographic.Chisq,2), as.character)) %>%
	select(Species:Environmental.corr.P, Geographic.Marginal.R2:Geographic.corr.P) %>%
	latex(file=curr.file, 
		cgroup=c('', 'Environmental', 'Geographic'), n.cgroup=c(1, 4, 4), col.just=rep('c', 9),
		digits=2, rowname=NULL, first.hline.double=FALSE, booktabs=TRUE, here=TRUE,
		colheads=c('\\bfseries{Species}', '$MR^2$', '$CR^2$', '$\\chi^2$', '\\textit{P}', '$MR^2$', '$CR^2$', '$\\chi^2$', '\\textit{P}'),
		caption='Summary of maximum likelihood population-effects (MLPE) models. Briefly, for each species associated with potentially adaptive loci, a model was fit to correlate dissimilarities among planning units where the species in terms of the adaptive genetic characteristics of individuals in them with dissimilarities in terms of the environmental conditions inside the planning units (denoted in the environmental column group). Similarly, for each species, a model was fit to correlate dissimilarities among the planning units in terms of the neutral genetic characteristics of indiviudals in them with dissimilarities in terms of the geographic position of the planning units (denoted in the geographic column group). Data shows the results from comparing each model to its corresponding null model using $\\mathbf{\\chi^2}$ tests (1 degree of freedom), and the $\\mathbf{R^2}$ ($\\mathbf{MR^2}$) and conditional $\\mathbf{R^2}$ ($\\mathbf{CR^2}$) for each model.'
	)
readLines(curr.file)[-1] %>% paste0(collapse='\n') %>% 
	gsub(pattern='\\cline{1-', replacement='\\cline{2-6} \\cline{8-', fixed=TRUE) %>% 
	gsub(pattern='{c}{\\textit{P}}\\tabularnewline', replacement='{c}{\\textit{P}}\\T\\B\\tabularnewline', fixed=TRUE) %>% 
	gsub(pattern='\\toprule', replacement='', fixed=TRUE) %>%
	gsub(pattern='\\midrule', replacement='\\midrule', fixed=TRUE) %>%
	gsub(pattern='\\bottomrule', replacement='\\midrule', fixed=TRUE) %>%
	cat()
```

<!-- surrogacy analysis -->
For each species, we generated a suite of `r raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates*2` random prioritisations, and calcuated the proportion of environmental variation and adaptive (if present) genetic variation they sampled. We then repeated this process, and calculated the proportion of geographic variation and neutral genetic variation they sampled. Environmental and geographic variables were moderately effective predictors for genetic variation for the genetic variation represented by the randomly generated prioritisations (Figure 1). The relationship between the proportion of genetic variation and surrogate variation secured in the prioritisations varied among species for environmental surrogates of adaptive genetic variation (species $\times$ proportion of environmental variation interaction term; ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$), and geographic surrogates of neutral variation (${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that the relationships for all species were significantly different to zero (environmental: minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S3). After establishing that the surrogate variables predicted how well prioritisations secured genetic variation, our final step was to determine if prioritisations were more effective when generated using surrogate-based targets.

@@

<!-- scenario analysis -->
To determine whether using environmental and geographic targets could improve the effectiveness of prioritisations in representing genetic variation, we generated prioritisations using (i) "amount targets" reflecting the traditional approach of representing a certain proportion of each species' goegraphic distribution, (ii) "amount and surrogate targets" where we also targeted representation of the environmental and geographic surrogates as well as a certain proportion of each species' geographic distribution, and (iii) "amount and genetic targets" where we targeted representation of the directly measured genetic variation as well as a certain proportion of each species' geographic distribution (Figure 3). To determine if these patterns were robust, for each of the three combinations of the targets, we generated prioritsations under four scenarios: (i) single-species with equal costs, (ii) single species with opportunity costs, (iii) multi-species with equal costs, and (iv) multi-species with opportunity costs.

@@

The proportion of genetic variation secured in a prioritisation depended on the targets used to generate the prioritisation (${\chi^2}_{`r scenario.ANOVA.4.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.4.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.4.DF$p.value[[2]])`)$. Most notably, prioritisations based on the surrogate variables represented significantly more genetic variation than amount-based prioritisations (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` overall genetic variation secured versus `r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)`; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[1]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[1]])`$), and were not distinguishable from prioritisations based on measured genetic data (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)`; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[3]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[3]])`$). These results indicate that environmental and geographic surrogates in this case far outperform traditional amount-based conservation planning, and perform as well as a conservation plan based directly on measured genetic data.

@@

<!--  introduction -->
We aimed to investigate the effectiveness of broad-scale environmental and geographic variables as surrogates for adaptive and neutral genetic variation in conservation planning. Our study is the first to address this aim using a diverse set of species and calibrated against field measurements of genetic variation. The spatial patterns of environmental and geographic surrogates was strongly correlated with the patterns of adaptive and neutral genetic variation for most species, and environmental and geographic targets resulted in prioritisations that secured a greater proportion of adaptive and neutral genetic variation than traditional conservation planning for most species. These environmental and geographic surrogates are based on freely available data, and could be calculated for any study region across the world.

<!-- surrogate/genetic dissimilarity matrix correlations -->
Our analyses showed that differences in the environmental conditions and geographic position among the planning units tended to correlate with genetic differences among the individuals in the planning units (Table 1). For both surrogates and their corresponding models, the conditional $R^2$ values (environmental vs. adaptive genetic variation: mean `r round(mean(correlation.model.summary.DF$Environmental.Conditional.R2,na.rm=TRUE),2)` conditional $R^2$ `r round(sd(correlation.model.summary.DF$Environmental.Conditional.R2,na.rm=TRUE),2)` $\pm$ SD; geographic vs. neutral genetic variation: mean `r round(mean(correlation.model.summary.DF$Geographic.Conditional.R2,na.rm=TRUE),2)` conditional $R^2$ `r round(sd(correlation.model.summary.DF$Geographic.Conditional.R2,na.rm=TRUE),2)` $\pm$ SD) were generally slightly higher than the marginal $R^2$ values (environmental vs. adaptive genetic variation: mean `r round(mean(correlation.model.summary.DF$Environmental.Marginal.R2,na.rm=TRUE),2)` marginal $R^2$ `r round(sd(correlation.model.summary.DF$Environmental.Marginal.R2,na.rm=TRUE),2)` $\pm$ SD; geographic vs. neutral genetic variation: mean `r round(mean(correlation.model.summary.DF$Geographic.Marginal.R2,na.rm=TRUE),2)` marginal $R^2$ `r round(sd(correlation.model.summary.DF$Geographic.Marginal.R2,na.rm=TRUE),2)` $\pm$ SD). This result suggests that while variation in the surrogate variables correlates with genetic variation, both forms of genetic variation contain additional spatially auto-correlated variation that is not explained by these surrogates. In particular, the differences between the conditional and marginal $R^2$ values for the models examining the effectiveness of geographic variables suggest that this unexplained spatial auto-correlation may be non-linear or driven by a few neighboring planning units with genetically similar individuals. Similarly, other studies have also reported strong patterns of spatial auto-correlation in patterns of genetic variation [eg. @r502].

>>>>

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Hugh P. Possingham
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  All authors developed the study and drafted the manuscript. JOH performed all analyses.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for processes to continue. However, evolutionary processes are not often explicitly considered when deciding where to site protected areas because genetic data require considerable resources to obtain and analyse. Here we show that environmental and geographic distance variables are effective surrogates for adaptive and neutral intra-specific variation for conservation planning. We obtained occurrence and genetic data from the IntraBioDiv project for 27 species collected over the European Alps using a gridded sampling scheme. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the grid cells. We then used the cells as planning units to generate prioritisations. Our first analysis verified that the spatial patterns of environmental and geographic variation correlated with those adaptive and neutral genetic variation (respectively). Our second analysis showed that these surrogates can predict the proportion of genetic variation secured in randomly generated solutions. Our third analysis found that solutions generated using additional surrogate-based targets secured more neutral genetic variation. We discovered that solutions generated using traditional amount-based targets tend to secure most species' adaptive genetic variation. Overall, our work provides a rigorous examination of two potential surrogates for genetic variation.
significance: |
  Protected areas are used to preserve biodiversity patterns and processes. Since the resources available for conservation are limited, protected areas need to be sited in cost-effective locations. However, despite their importance, evolutionary processes are not often used to guide reserve selection. This is primarily because genetic data require extensive resources to obtain and process. Here we show that freely available environmental and geographic variables can be used as surrogates for genetic data in conservation planning.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Biodiversity and Conservation Science.
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

<!-- Introduction -->
Protected areas spearhead conservation efforts [@r440]. They buffer species from anthropogenic impacts; providing places for them to persist. Since the resources available for conservation are limited, protected areas need to be sited in places that fulfill conservation objectives for minimal cost [@r255]. To achieve this, conservation planning exercises often generate plans for entire networks of protected areas (prioritisations) to allocate resources cost-effectively and preserve broad-scale biodiversity processes. Intra-specific genetic variation is fundamental to long-term species persistence [@r165], and as a consequence, there has been increasing interest in designing prioritisations that capture and cultivate this aspect of biodiversity [@r67; @r3; @r61; @r499].

Broadly speaking, though the strength of natural selection is a continuous force, genetic variation can be classified as either adaptive or neutral [reviewed in @r480]. Adaptive genetic variation is associated with loci that significantly affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals [eg. @r456; @r487], though phenotypic-genetic associations are more likely to identify functionally important genomic regions for individual species [eg. @r500]. By representing the full range of adaptive variation, protected areas can help ensure that populations with particularly beneficial adaptations are not lost [@r3; @r43]. In contrast, neutral genetic variation is associated with loci that do not significantly affect fitness, but instead reflects the evolutionary history of different populations. By representing the full range of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity [@r67]. Thus optimally sited protected area networks would aim to represent both adaptive and neutral patterns of genetic variation [@r67; @r3; @r43].

Recently, the use of surrogate data has been proposed to generate prioritisations that sufficiently intra-specific genetic variation--without needing to directly utilise genetic data to achieve this. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been proposed to guide reserve selection as surrogates for adaptive genetic variation [eg. @r202; @r2]. By generating prioritisations that preserve a more representative sample of environmental conditions across species' ranges, these prioritisations could also capture more adaptive variation. However, the effectiveness of environmental surrogates in representing adaptive genetic variation remains unverified. Neutral genetic variation arises from a reduction in gene flow between individuals. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood [@r501]. Potential surrogates for neutral variation have been based on variables that predict the levels of connectivity between planning units. For instance, isolation by distance [IBD; @r259] predicts that populations further apart will experience less gene flow, and in turn, share less neutral loci. Based on this idea, geographic surrogates based on distance have been found to improve conservation prioritisations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. habitat type @r497]--the typical situations in most conservation planning exercises.

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods to detect loci under selection, and they returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD; Table S1). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions was needed to sufficiently describe their patterns of adaptive (all $K = `r unique(na.omit(adaptive.k))`$; Table S2) and neutral genetic variation ($K = `r min(neutral.k)`$--$`r max(neutral.k)`$; Table S1). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S29--S54) generally showed substantial spatial auto-correlation (eg. _Cerastium uniflorum_, Figure S34; _Dryas octopetala_, Figures S36), suggesting that the environmental and geographic variables may prove to be effective surrogates.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental and geographic variation correlated with the broad-scale spatial patterns in adaptive and neutral variation for most species (respectively). For each species,we constructed dissimilarity matrices based on the planning units they were found in. These matrices expressed differences between the planning units in terms of their (i) environmental characteristics, (ii) geographic position, and also the (iii) adaptive (if detected), and (iv) neutral genetic characteristics of individuals found inside them. By correlating these matrices, the spatial patterns of environmental variation were found to be significantly correlated with that of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models; Table S3). Similarly, geographic distance between planning units was also correlated significantly with the spatial pattern of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models; Table S3). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these patterns were correlated, our next step was to determine whether they were correlated enough to be effective for conservation planning.

<!-- surrogacy analysis -->
For each species, we generated a suite of `r format(raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates, big.mark=',', scientific=FALSE)` random prioritisations, and calculated the proportion of environmental variation and adaptive (if present) genetic variation they sampled. We then repeated this process, and calculated the proportion of geographic variation and neutral genetic variation they sampled. The environmental and geographic variables were moderately effective predictors for the genetic variation represented by the randomly generated prioritisations (Figure 1). The relationship between the proportion of genetic and surrogate variation secured in the prioritisations varied among the species (species $\times$ proportion of environmental variation interaction term: ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$; species $\times$ proportion of geographic variation interaction term: ${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that these relationships were positive for all species (environmental: minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S4). After establishing that the the proportion of surrogate variation secured in a prioritisation also predicted the proportion of genetic variation it secured, our final step was to determine if prioritisations were more effective when generated using targets based on the surrogates.

@@

<!-- scenario analysis -->
To determine whether using environmental and geographic targets could improve the effectiveness of prioritisations in representing genetic variation, we generated prioritisations using (i) "amount targets" reflecting the traditional approach of representing a certain proportion of each species' geographic distribution, (ii) "amount and surrogate targets" where we also targeted representation of the environmental and geographic surrogates as well as a certain proportion of each species' geographic distribution, and (iii) "amount and genetic targets" where we targeted representation of the directly measured genetic variation as well as a certain proportion of each species' geographic distribution (Figure 3). To determine if these patterns were robust, for each of the three combinations of the targets, we generated prioritisations under four scenarios: (i) single-species with equal costs, (ii) single species with opportunity costs, (iii) multi-species with equal costs, and (iv) multi-species with opportunity costs.

@@

The proportion of genetic variation secured in a prioritisation depended on the targets used to generate the prioritisation (${\chi^2}_{`r scenario.ANOVA.4.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.4.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.4.DF$p.value[[2]])`)$. Most notably, prioritisations based on the surrogate variables represented significantly more genetic variation than amount-based prioritisations (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` SD overall genetic variation secured versus `r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[1]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[1]])`$), and were not distinguishable from prioritisations based on measured genetic data (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[3]],2)`$; $P = `r pvalString(scenario.GLHT.2.DF$p.value[[3]])`$). These results indicate that environmental and geographic surrogates in this case far outperform traditional amount-based conservation planning, and perform as well as a conservation plan based directly on measured genetic data.

@@

<!--  introduction -->
We aimed to investigate the effectiveness of broad-scale environmental and geographic variables as surrogates for adaptive and neutral genetic variation in conservation planning. Our study is the first to address this aim using a broad range of species and calibrated against field measurements of genetic variation. The spatial patterns of environmental and geographic surrogates were strongly correlated with the patterns of adaptive and neutral genetic variation for most species, and environmental and geographic targets resulted in prioritisations that secured a greater proportion of adaptive and neutral genetic variation than traditional conservation planning for most species. These environmental and geographic surrogates are based on freely available data, and could be calculated for any study region across the world.

<!-- surrogate/genetic held correlations -->
Our results demonstrate that environmental and geographic surrogates are strong enough to be operational (Figure 2). For most species, the environmental and geographic surrogates variables were effective surrogates for adaptive (mean `r round(mean(env.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(env.surrogate.spp.DF$r2CU),2)` SD; Figures 2c and S56) and neutral genetic variation (mean `r round(mean(geo.surrogate.spp.DF$r2CU),2)` $R^2$ $\pm$ `r round(sd(geo.surrogate.spp.DF$r2CU),2)` SD; Figures 2d and Figure S57). Despite this, for a few species, geographic distance was a poor surrogate for neutral genetic variation (eg. _Gypsophila repens_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gypsophila_repens')$Geographic.r2CU,2)`; _Luzula alpinopilosa_: $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Luzula_alpinopilosa')$Geographic.r2CU,2)`; _Gentiana nivalis_ $R^2$ = `r round(filter(surrogate.spp.DF, Species=='Gentiana_nivalis')$Geographic.r2CU,2)`). Indeed, geographic distances will be a poor surrogate for species that utilise dispersal modes that result in non-linear relationships between connectivity and geographic distance [eg. wind dispersing plants can maintain high connectivity between distant areas; @r451], and in places where connectivity is most affected by landscape features [@r497]. Overall, however, such species were the exception in our analysis, and both simple surrogates performed moderately well for a diverse set of species.
